\bf An Examination of Sport Climbing Competition Format and Scoring System

### 2021-08-04 manuscript/Rmd/draft.Rmd
Trying to finish

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world (CITATION NEEDED - I mentioned the article at the end of the paragrah). In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint" (CITATION NEEDED). She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

@@

To the best of our knowledge, we know of no sporting event, team or individual, that uses the product of ranks to determine an overall rankings.  There are examples of team sports that use the sum of ranks to determine the winning team such as cross country, where the team with the lowest sum of ranks of the top five runners is determined to be the winner. For more detail on rank sum scoring, see @hammond2007 and @boudreau2018.

Some individual sports such as the decathlon and heptathlon rely on a sum of scores from the ten or seven events, however, these scores are not determined based on the ranks of the competitors.  That is a decathlete's score is entirely based upon their times, distances, and heights, and their overall score will be exactly the same if the times, distances, and heights remain the same regardless of the performance of other individuals.  Full details of decathlon scoring can be found in @westera2006.

There are other individual competitions consisted of several events combined that do base their scoring on ranks, such as crossfit competitions.  In each event, points are earned based on the competitor's rank in the event based on a scoring table. (https://games.crossfit.com/announcement/12940), and a competitors final score is based on the sum of their scores across all the events.

@@

As a side note, rank product, genes

@@

```{r}
set.seed(1)
nsim <- 10000
nplay <- 20
qual_sim <- list()
for (i in 1:nsim) {
  qual_sim[[i]] <-
    bind_cols(
      player = 1:nplay,
      e1 = sample(1:nplay, replace = FALSE),
      e2 = sample(1:nplay, replace = FALSE),
      e3 = sample(1:nplay, replace = FALSE)
    ) %>%
    mutate(sim = i)
}

qual <- bind_rows(qual_sim) %>% 
  mutate(score = e1 * e2 * e3) %>% 
  group_by(sim) %>% 
  mutate(rank = rank(score, ties.method = "random")) %>%
  ungroup()
```

```{r}
set.seed(1)
nsim <- 10000
nplay <- 8
final_sim <- list()
for (i in 1:nsim) {
  final_sim[[i]] <-
    bind_cols(
      player = 1:nplay,
      e1 = sample(1:nplay, replace = FALSE),
      e2 = sample(1:nplay, replace = FALSE),
      e3 = sample(1:nplay, replace = FALSE)
    ) %>%
    mutate(sim = i)
}

final <- bind_rows(final_sim)

final <- final %>% 
  mutate(score = e1 * e2 * e3) %>% 
  group_by(sim) %>% 
  mutate(rank = rank(score, ties.method = "random")) %>%
  ungroup()

qual_final <- qual %>% 
  mutate(round = "Qualification") %>% 
  bind_rows(mutate(final, round = "Final")) %>% 
  mutate(round = fct_rev(round))
```

>>>>

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world (CITATION NEEDED). In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint" (CITATION NEEDED). She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

@@

To the best of our knowledge, we know of no sporting event, team or individual, that uses the product of ranks to determine an overall rankings.  There are examples of team sports that use the sum of ranks to determine the winning team such as cross country, where the team with the lowest sum of ranks of the top five runners is determined to be the winner. For more detail on rank sum scoring, see @hammond2007 and @boudreau2018. In addition, some individual sports such as the decathlon and heptathlon rely on a sum of scores from the ten or seven events, however, these scores are not determined based on the ranks of the competitors.  That is, a decathlete's score is entirely based upon their times, distances, and heights, and their overall score will be exactly the same if the times, distances, and heights remain the same regardless of the performance of other individuals.  Full details of decathlon scoring can be found in @westera2006.

<!-- 
here are other individual competitions consisted of several events combined that do base their scoring on ranks, such as crossfit competitions.  In each event, points are earned based on the competitor's rank in the event based on a scoring table. (https://games.crossfit.com/announcement/12940), and a competitors final score is based on the sum of their scores across all the events.

@@

As a side note, rank product, genes 
-->

@@

```{r}
climbing_sim <- function(nsim = 10000, nplay) {
  sims <- list()
  for (i in 1:nsim) {
    sims[[i]] <-
      bind_cols(
        player = 1:nplay,
        e1 = sample(1:nplay, replace = FALSE),
        e2 = sample(1:nplay, replace = FALSE),
        e3 = sample(1:nplay, replace = FALSE)
      ) %>%
      mutate(sim = i)
  }
  results <- bind_rows(sims) %>% 
    mutate(score = e1 * e2 * e3) %>% 
    group_by(sim) %>% 
    mutate(rank = rank(score, ties.method = "random")) %>%
    ungroup()

  return(results)
}

set.seed(1)
qual <- climbing_sim(nsim = 10000, nplay = 20)
final <- climbing_sim(nsim = 10000, nplay = 8)
```

```{r}
qual_final <- qual %>% 
  mutate(round = "Qualification") %>% 
  bind_rows(mutate(final, round = "Final")) %>% 
  mutate(round = fct_rev(round))
```

### 2021-08-04 manuscript/Rmd/draft.tex
Trying to finish

The decision to combine the three climbing events and only award one set
of medals for both men's and women's events in the Olympics has received
a large amount of criticism from climbing athletes all over the world
(CITATION NEEDED - I mentioned the article at the end of the paragrah).
In a series of interviews conducted by Climbing Magazine in 2016, a
number of climbers shared their thoughts and concerns about the new
Olympics climbing format. Climber Lynn Hill compared the idea of
combining speed climbing, bouldering, and lead climbing to ``asking a
middle distance runner to compete in the sprint'' (CITATION NEEDED). She
then added ``Speed climbing is a sport within our sport''. Other
climbers also hold the same opinion as Hill regarding speed climbing,
using words and phrases like ``bogus'', ``a bummer'', ``less than
ideal'', ``not in support'', and ``cheesy and unfair'' to describe the
new combined competition format. Courtney Woods stated ``Speed climbers
will have the biggest disadvantage because their realm isn't based on
difficult movements''. Mike Doyle believed ``Honestly, the people that
will suffer the most are the ones that focus only on speed climbing.
Those skills/abilities don't transfer as well to the other
disciplines''. Most climbers also expressed their hope for a change in
the format in future tournaments, with some calling for each discipline
to have its own set of medals. See \citet{blanchard2016} for more
information on these interviews.

@@

To the best of our knowledge, we know of no sporting event, team or
individual, that uses the product of ranks to determine an overall
rankings. There are examples of team sports that use the sum of ranks to
determine the winning team such as cross country, where the team with
the lowest sum of ranks of the top five runners is determined to be the
winner. For more detail on rank sum scoring, see \citet{hammond2007} and
\citet{boudreau2018}.

Some individual sports such as the decathlon and heptathlon rely on a
sum of scores from the ten or seven events, however, these scores are
not determined based on the ranks of the competitors. That is a
decathlete's score is entirely based upon their times, distances, and
heights, and their overall score will be exactly the same if the times,
distances, and heights remain the same regardless of the performance of
other individuals. Full details of decathlon scoring can be found in
\citet{westera2006}.

There are other individual competitions consisted of several events
combined that do base their scoring on ranks, such as crossfit
competitions. In each event, points are earned based on the competitor's
rank in the event based on a scoring table.
(\url{https://games.crossfit.com/announcement/12940}), and a competitors
final score is based on the sum of their scores across all the events.

Paragraph on scoring methods in other sports: Soccer points:
\citet{haugen2008}; Cross country scoring: \citet{hammond2007},
\citet{boudreau2018}; Decathlon scoring: \citet{westera2006}

As a side note, rank product, genes

@@

\caption{\label{tab:unnamed-chunk-4}Descriptive statistics of simulated scores for qualification and final rounds}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
round & min & Q1 & median & Q3 & max & mean & sd & n\\
\midrule
Qualification & 1 & 240 & 684 & 1638 & 8000 & 1158.84378 & 1273.47884 & 200000\\
Final & 1 & 24 & 60 & 126 & 512 & 91.12077 & 91.35293 & 80000\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:unnamed-chunk-7}The average scores for all final ranks according to our simulations. Ranks 1, 2, and 3 are table positions that guarantee medalist status for climbers.}
\centering
\begin{tabular}[t]{rr}
\toprule
Rank & Average score\\
\midrule
1 & 9.6748\\
2 & 20.4143\\
3 & 33.2648\\
4 & 50.5933\\
5 & 74.7620\\
\addlinespace
6 & 110.0536\\
7 & 164.4274\\
8 & 265.7760\\
\bottomrule
\end{tabular}
\end{table}

>>>>

The decision to combine the three climbing events and only award one set
of medals for both men's and women's events in the Olympics has received
a large amount of criticism from climbing athletes all over the world
(CITATION NEEDED). In a series of interviews conducted by Climbing
Magazine in 2016, a number of climbers shared their thoughts and
concerns about the new Olympics climbing format. Climber Lynn Hill
compared the idea of combining speed climbing, bouldering, and lead
climbing to ``asking a middle distance runner to compete in the sprint''
(CITATION NEEDED). She then added ``Speed climbing is a sport within our
sport''. Other climbers also hold the same opinion as Hill regarding
speed climbing, using words and phrases like ``bogus'', ``a bummer'',
``less than ideal'', ``not in support'', and ``cheesy and unfair'' to
describe the new combined competition format. Courtney Woods stated
``Speed climbers will have the biggest disadvantage because their realm
isn't based on difficult movements''. Mike Doyle believed ``Honestly,
the people that will suffer the most are the ones that focus only on
speed climbing. Those skills/abilities don't transfer as well to the
other disciplines''. Most climbers also expressed their hope for a
change in the format in future tournaments, with some calling for each
discipline to have its own set of medals. See \citet{blanchard2016} for
more information on these interviews.

@@

To the best of our knowledge, we know of no sporting event, team or
individual, that uses the product of ranks to determine an overall
rankings. There are examples of team sports that use the sum of ranks to
determine the winning team such as cross country, where the team with
the lowest sum of ranks of the top five runners is determined to be the
winner. For more detail on rank sum scoring, see \citet{hammond2007} and
\citet{boudreau2018}. In addition, some individual sports such as the
decathlon and heptathlon rely on a sum of scores from the ten or seven
events, however, these scores are not determined based on the ranks of
the competitors. That is, a decathlete's score is entirely based upon
their times, distances, and heights, and their overall score will be
exactly the same if the times, distances, and heights remain the same
regardless of the performance of other individuals. Full details of
decathlon scoring can be found in \citet{westera2006}.

@@

\caption{\label{tab:unnamed-chunk-4}Descriptive statistics of simulated scores for qualification and final rounds}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
round & min & Q1 & median & Q3 & max & mean & sd & n\\
\midrule
Qualification & 1 & 240 & 684 & 1638 & 8000 & 1158.84378 & 1273.47884 & 200000\\
Final & 1 & 24 & 60 & 126 & 512 & 91.19935 & 91.11065 & 80000\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:unnamed-chunk-7}The average scores for all final ranks according to our simulations. Ranks 1, 2, and 3 are table positions that guarantee medalist status for climbers.}
\centering
\begin{tabular}[t]{rr}
\toprule
Rank & Average score\\
\midrule
1 & 9.6687\\
2 & 20.1844\\
3 & 33.3658\\
4 & 50.5734\\
5 & 75.2499\\
\addlinespace
6 & 110.7070\\
7 & 164.8258\\
8 & 265.0198\\
\bottomrule
\end{tabular}
\end{table}
